How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2024-09-11 00-author.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

This book is written in bookdown [@xie2016bookdown] specifically to make it a 'live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book. The easiest way for readers to contribute content directly is through a [GitHub pull requests](https://help.github.com/articles/about-pull-requests/). At the repository for this book, you will find Rmd files for each chapter, and as a GitHub user, you can simply edit the Rmd file and submit the changes. If I am happy with the changes proposed, I will merge your content with that of the book and add your name to the [Acknowledgements](#acknowledge).

One of the amazing potentials for [bookdown books](https://bookdown.org/) is that all the files for this book are hosted in a repository on [Github](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/tree/main). You have the opportunity to fork this repository and write your own version for a different discipline, a different language or for a different region of the world. It is also my hope that this guide can grow to become a community of practice for those conducting PhDs in Biological Sciences. It will not be possible to cover every aspect of writing a PhD in Biological Sciences, but it may be that I have missed out ones that are very important to you. Equally, parts of what is currently written will become obsolete as new initiatives begin, and old problems are resolved. For this reason, this guide needs to be a 'living document', and anyone who wants to provide feedback or contribute new sections is more than welcome. Please feel free to open an issue, or make a Pull Request if you spot a typo.

## How to cite {-}
Measey, J. (2022) *How to write a PhD in Biological Sciences: a guide for the uninitiated*. CRC Press, Boca Raton. Online Version: **`r Sys.Date()`**  ISBN: 9781032080208

>>>>

This book is written in bookdown [@xie2016bookdown] specifically to make it a 'live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book. The easiest way for readers to contribute content directly is through a [GitHub pull request](https://help.github.com/articles/about-pull-requests/). At the repository for this book, you will find Rmd files for each chapter, and as a [GitHub user](https://github.com/signup?ref_cta=Sign+up&ref_loc=header+logged+out&ref_page=%2F&source=header-home), you can simply edit the Rmd file and submit the changes. If I am happy with the changes proposed, I will merge your content with that of the book and add your name to the [Acknowledgements](#acknowledge).

One of the amazing potentials for [bookdown books](https://bookdown.org/) is that all the files for this book are hosted in a repository on [Github](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/tree/main). You have the opportunity to fork this repository and write your own version for a different discipline, a different language, or for a different region of the world. It is also my hope that this guide can grow to become a community of practice for those conducting PhDs in Biological Sciences. It will not be possible to cover every aspect of writing a PhD in Biological Sciences, and it may be that I have missed out ones that are very important to you. Equally, parts of what is currently written will become obsolete as new initiatives begin, and old problems are resolved. For this reason, this guide needs to be a 'living document', and anyone who wants to provide feedback or contribute new sections is more than welcome. Please feel free to open an issue, or make a Pull Request if you spot an error.

## How to cite {-}
Measey, J. (2022) *How to write a PhD in Biological Sciences: a guide for the uninitiated*. CRC Press, Boca Raton.  262pp ISBN: 9781032080208  DOI:10.1201/9781003212560

Measey, J. (`r format(Sys.time(), '%Y')`) *How to write a PhD in Biological Sciences: a guide for the uninitiated*. Online Version from https://howtowriteaphd.org/: **`r format(Sys.time(), '%d %B, %Y')`**

[Download Bibtex file here](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/blob/main/citation.bib)

### 2024-09-11 1.00-Part1.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you plan to do afterwards. There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make informed decisions from the outset. For example, reconciling yourself to using the [statistical package R](#useR) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on. Preparing yourself into the right mental framework will also be very important for your ability to work under pressure. Getting into good habits at the beginning will put you in a commanding position later. The aim of this first part of the book is to help you assess your current position, and realign yourself toward an optimum mindset.

>>>>

Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you plan to do afterwards. Both short-term and long-term planning will be one of the [core skills](#coreskills) that are the subject of Part I. You are likely to have had experience of each of these core skills before, but this part of the book tries to show you the importance of growing these skills and having the discipline to apply them regularly to achieve your goal of finishing  your PhD. Some of the most important core skills involve critical reading and regular writing, skills that you will need to begin acquiring and practising from the outset of your studies. Managing your time day-to-day will help you find time to practice these skills alongside your other commitments. There are a lot of things to think about as you start your PhD, and for many of these, it is far easier to make informed decisions from the outset. Preparing yourself for the right mental state will also be very important for your ability to work under pressure. Getting into good habits at the beginning will put you in a commanding position later. This first part of the book aims to help you assess your current position, and realign yourself toward an optimum mindset for doing your PhD in Biological Sciences.

### 2024-09-11 1.01-beforeyoustart.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

# So you are doing a PhD? {#beforeyoustart}

Universities are worried about two major aspects of those who start PhD studies:

 - The time to completion in many countries worldwide is much longer than it should be. For example, 7.5 years is the median time to completion in the USA [@geven2018how], Europe universities have largely adopted the Salzburg Principles that doctoral programmes should be three to four years for full-time candidates  [@christensen2005bologna]. Following this, an average is recorded as being between 3.5 and 4.5 years, with most institutions having reduced this time over the last 10 years [@hasgall2019doctoral].
 - Completion rates are shockingly low. As bad as 50% in the USA [@litalien2015improving], 66% (within 6 years) in Europe [@hasgall2019doctoral], but generally better in the sciences [@park2005new].

This has spurred many responsible for postgraduate students to try to ensure that they receive more support for their studies. But there is also the need for students to comprehend right from the start that they are undertaking a major commitment that will likely impact many different parts of their lives.

In this first chapter of the book, I intend to make you familiar with what a PhD is, how to approach thinking about it, and what to look for in an institution and advisor. You may already have some or all of these things fixed, but there is still some use in going over these basic points. Try to keep an open mind. Doing a PhD will take several years of your life, and may well define your future career prospects. Therefore, to say that this is a big decision is an understatement. It is not a substitute for getting a job or a way of remaining a student for longer. If after reading this section you are in any doubt about whether or not a PhD is right for you, then I would suggest that you do need to think about it further. Don't take my word for it, go and talk to people who have done a PhD (whether or not they now have a job), or those who are doing a PhD now. Especially if their experience is in your focal research area, they are likely to know a lot more about it.

@@

@@

- First is the clear ability to conceptualise and carry out a body of scientific work that is novel. Here I mean novel in that it has not simply been a repeat of some previous work on a different taxon or system. This is easier to judge from an oral defense, over reading the thesis. 
- Second is the clear knowledge of how the work that is conducted fits into the bigger picture. That is, that the work demonstrates that the person is a scientist at a broader level than the narrow focus of their research.
- Third is whether the student has been able to bring the study together in a way that shows that it is more than the sum of its parts. In essence, together the 5 data chapters should provide a bigger overview of a broader topic, and show the direction of future work.

@@

## What will you do after your PhD?
It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most of what you are doing now. During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas. Knowing what you want to do when you finish will help guide your decisions so that you can maximise opportunities as they arise. Both inside and outside of academia, it is likely to be the contacts that you make that will help in your next career move.

### Doing a PhD actually reduces your chance of employment and lowers your pay
The reality is that most employers don’t want someone with a PhD, because it is a very academic qualification. It will take you so far down an academic rabbit hole, that for an employer there’s very little to use. Most employers want people who understand the nuts and bolts of doing research, but who can apply this knowledge outside of the academic system. Employers who want people with PhDs (and there are a few) generally want professional researchers who are essentially academics. Good work if you can get it, but it tends not to pay well. You might end up only being fit to be an academic, and that life is getting increasingly harder without any end to the expectations of the employers… like the equivalent of doing a PhD every year. Something only very few people will enjoy.

At the same time, governments the world over, and especially developing countries, are providing bursaries for people to do PhDs. Why? The number of PhDs in a country is considered  (by the [World Bank](https://datacatalog.worldbank.org/dataset/world-development-indicators)) to be one of many measures of the development of that country. As the percentage gets larger, so countries can convince the World Bank that they are developing (and then get granted its favours, see Figure \@ref(fig:PhDMap)). This is all well and good, but strategically, governments should consider why they want these researchers in their populations, and more particularly for what kinds of jobs. It seems somewhat unfair to produce lots of young PhD holders that have no careers to join, or prospects to use their PhD just so that your government can tick a check box on their road to development, and so borrow more money from the World Bank. If you are from a developing country, you cannot assume that just because you are given a bursary by your government it follows that they will give you a job or a career. This will likely always be your own task.

(ref:PhDMap) **Where in the world are there most PhDs?** In this map of the world, countries shaded are those that have data declaring the proportion of PhD (or equivalent) in their population (max percentage between 2010 and 2020 (Data from the World Bank). This ranges from 2.97% in Switzerland (dark red) to 0.02% in DR Congo (light yellow). Notably, there is no data held in these years for many countries in Asia, Africa and South America.

@@

In a recent report from the UK, it was found that by age 35, the earnings gap for those with PhDs was lower than the same cohort that left academia after their BSc or MSc [@britton2020earnings]. Women who obtain PhDs tend to increase their earnings by ~8%, while men _reduce_ their earnings by ~9%. However, for those in Biological Sciences women seem likely to earn no more than those with undergraduate degrees by age 35. Men earn around 10% less. The point here is that you should not be doing a PhD in Biological Sciences because you think you will earn more money in your following career. Having made this point, I also feel that it is important to point out that how much you earn may not be directly related to job satisfaction or lifetime happiness. To my knowledge, this is not (yet) reported, but could be a lot higher for those with PhDs, even though they earn less.

It is important to realise that many academics are looking for good students to do PhD work. This is because they need someone to do research within their groups, and students or postdocs do most of this work [see @chamberlain2016ten]. For this reason, an academic you know while you're doing your MSc studies may start to try to persuade you to do a PhD. This could be fantastic for you, it could be just what you're looking for. But the reality is that you will never do a PhD for somebody else, you are always going to be doing it for yourself. If the only reason you are doing a PhD is because someone else told you that you should then you are likely to have a really hard time. By the time you graduate, you may have a PhD but without any idea of what you're going to do with it.

@@

@@

## Your mental health will likely suffer
Doing a PhD will be the cause of stress not only in your working life, but will also impact other areas of your life. Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a support network. It is worth tracking your mental health during your PhD to check that you are not getting into difficulties, and I show you later [how to do this](#healthy).

@@

## What to avoid in an advisor
Remember that relationships occur on a one-to-one basis, and it may be that not every advisor gets on with every student. There are some advisors who have many problems with many students, and these are best avoided no matter how high their profile is [see @malaga2004meyer;@abbott2004junior; but also see @vences2004meyer for alternative viewpoints]. There are other advisors who are so busy that they are never able to meet with their students, and pass this task on to somebody else in their group. Again no matter how high profile these people are, if they can't advise their own students they are best avoided. Chamberlain [-@chamberlain2016ten] suggests that there are 10 types of academic advisors, implying that 9 of these are rather disfunctional.

@@

It is possible that your potential advisor will tell you that they don't have any funding for a project that you may want to do. This is both a curse and an opportunity. There are many opportunities to raise money to fund PhD projects. Raising the funding yourself is one way of making sure it really remains your project. But the availability of this kind of funding does vary from country to country and you may want to embark on the fundraising side of your PhD before you register. Especially if you have a limited amount of time once you have registered in order to complete your PhD studies. 
Some projects are much cheaper than others. For example, if your intention is to sequence genomes of every animal or plant that you work with, the finance required will likely be far greater than your bursary. If however your intention is to do a PhD that models distributions of populations of animals and plants, then the entire costs for your study may only add up to a computer, remote access to a computing cluster and data storage.

@@

>>>>

# From the start of your PhD {#beforeyoustart}

The start of your PhD studies are a very turbulent time. You may be in a new institution with new rules to follow. You may be far away from the people at home and friends who make up your [support network](#support). You will be meeting many new people and finding lots of other people who are interested in the same subjects as you. You should be talking about your shared ideas and aspirations for your PhD studies. Although it is a turbulent time, it is also a very exciting time.

In this first chapter of the book, I intend to make you familiar with what a PhD is, how to approach thinking about it, and what to look for in an institution and advisor. You may already have some or all of these things fixed, but there is still some use in going over these basic points. Try to keep an open mind. Doing a PhD will take several years of your life, and may well define your future career prospects. Therefore, to say that this is a big decision is an understatement. It should not be a substitute for getting a job or a way of remaining a student for longer. If after reading this section you are in any doubt about whether or not a PhD is right for you, then I would suggest that you do need to think about it further. Don't take my word for it, go and talk to people who have done a PhD (whether or not they now have a job), or those who are doing a PhD now. Especially if their experience is in your focal research area, they are likely to know a lot more about it.

Your advisor and your institution are also happy to have you start your PhD. Your research and findings will be important to them. But they will also want to make sure that you finish your studies in a timely manner, as too many students start but do not finish their work, and many take too long to complete their studies. Universities are worried about two major aspects of those who start PhD studies:

 - **The time to completion** in many countries worldwide is much longer than it should be. For example, 7.5 years is the median time to completion in the USA [@geven2018how], Europe universities have largely adopted the Salzburg Principles that doctoral programmes should be three to four years for full-time candidates  [@christensen2005bologna]. Following this, an average is recorded as being between 3.5 and 4.5 years, with most institutions having reduced this time over the last 10 years [@hasgall2019doctoral]. Similarly, in China doctoral studies take three to four years to complete, although there are no official available statistics on time to completion [@powell2007doctorate; @zhu2024graduate]. 
 - **Completion rates** are shockingly low. As bad as 50% in the USA [@litalien2015improving], 66% (within 6 years) in Europe [@hasgall2019doctoral], but generally better in the sciences than other subjects [@park2005new]. Once again, there appear to be no official statistics for completion rates for those enrolled in Chinese PhD programmes [@zhu2024graduate].

This has spurred many responsible for postgraduate students to try to ensure that they receive more support for their studies. But there is also the need for you as the student to comprehend right from the start that you are undertaking a major commitment that will likely impact many different parts of their lives.

@@

It may be worth briefly considering the history of this qualification. Fifty years ago, a PhD was only taken by a small number of elite academics for whom it was treated as an apprenticeship to themselves become an academic [@houston2024final]. This has now changed, and in the 21st century, a PhD has become a qualification that can also be sought after outside academia. In some countries, where industry has an important voice in the content of a PhD, the synergy between industry and education appears to lead to national prosperity [@frick2017death]. This, in turn, leads to important questions about what a PhD is and who it is for, especially when one considers the use of PhD statistics by the World Bank (see [below](#lowpay)).

@@

- First is the clear ability to conceptualise and carry out a body of scientific work that is novel. Here I mean novel in that it has not simply been a repeat of some previous work on a different taxon or system. This is easier to judge from an oral defense, over reading the thesis. 
- Second is the clear knowledge of how the work that is conducted fits into the bigger picture. That is, that the work demonstrates that the person is a scientist at a broader level than the narrow focus of their research.

- Third is whether the student has been able to bring the study together in a way that shows that it is more than the sum of its parts. In essence, together the 5 data chapters should provide a bigger overview of a broader topic, and show the direction of future work.

@@

## What will you do after your PhD?
It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most of what you are doing now (see more on planning [here](#Plan)). During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas. Knowing what you want to do when you finish will help guide your decisions so that you can maximise opportunities as they arise. Both inside and outside of academia, it is likely to be the contacts that you make that will help in your next career move.

### Doing a PhD actually reduces your chance of employment and lowers your pay {#lowpay}
The reality is that most employers don’t want someone with a PhD, because it is a very academic qualification. It will take you so far down an academic rabbit hole, that for an employer there’s very little to use. PhDs were originally designed by academics as an apprenticiship for future academics [@houston2024final]. Most employers want people who understand the nuts and bolts of doing research, but who can apply this knowledge outside of the academic system. Employers who want people with PhDs (and there are a few) generally want professional researchers who are essentially academics. Good work if you can get it. You might end up only being fit to be an academic (a traditionally low paid sector which is likely what skews the figures of salaries of PhD holders), and that life is getting increasingly harder without any end to the expectations of the employers. Something only very few people will enjoy.

At the same time, governments the world over, and especially developing countries, are providing bursaries for people to do PhDs. Why? The number of PhDs in a country is considered  (by the [World Bank](https://datacatalog.worldbank.org/dataset/world-development-indicators)) to be one of many measures of the development of that country. This is also true in East Asia, where the dramatic increase of postgraduate degrees in the last two decades has primarily been driven by efforts to accelerate national economic development [@jung2022expansion]. As the percentage of PhD holders per 100 000 of population gets larger, so countries can convince the World Bank that they are developing (and then get granted its favours, see Figure \@ref(fig:PhDMap)). This is all well and good, but strategically, governments should consider why they want these researchers in their populations, and more particularly for what kinds of jobs. It seems somewhat unfair to produce lots of young PhD holders that have no careers to join, or prospects to use their PhD just so that your government can tick a check box on their road to development, and so borrow more money from the World Bank. If you are from a developing country, you cannot assume that just because you are given a bursary by your government it follows that they will give you a job or a career. This will likely always be your own task.

(ref:PhDMap) **Where in the world are there most PhDs?** In this map of the world, countries shaded are those that have data declaring the proportion of PhD (or equivalent) in their population (max percentage between 2010 and 2024 (Data from the [World Bank](https://databank.worldbank.org/reports.aspx?source=Education%20Statistics)). This ranges from 3.138% in Switzerland (red) to 0.008% in Guatemala (light yellow). Notably, there is no data (white) held in these years for some countries.

@@

In a recent report from the UK, it was found that by age 35, the earnings gap for those with PhDs was lower than the same cohort that left academia after their BSc or MSc [@britton2020earnings]. Women who obtain PhDs tend to increase their earnings by ~8%, while men _reduce_ their earnings by ~9%. However, for those in Biological Sciences women seem likely to earn no more than those with undergraduate degrees by age 35. Men earn around 10% less. Studies with data from other countries are broadly in agreement [@polonyi2022it]. The point here is that you should not be doing a PhD in Biological Sciences because you think you will earn more money in your following career. Similarly, another qualitative study of the perception of PhD holders in the UK suggests that while doing a PhD provides a lot of personal value for the student, it lacks any monetary value from the career that follows: i.e. their PhD Worth doing but not worth having [@guccione2022worth]. Another way of thinking about this is that it might be more important to you to have job satisfaction or lifetime happiness over a well paid job. To my knowledge, comparative levels of happiness is not (yet) reported, but could be a lot higher for those with PhDs, even though they may earn less.

It is important to realise that many academics are looking for good students to do PhD work. This is because they need someone to do research within their groups, and students or postdocs do most of this work [see @chamberlain2016ten]. For this reason, an academic you know while you're doing your MSc studies may start to try to persuade you to do a PhD. This could be fantastic for you, it could be just what you're looking for. But the reality is that you will never do a PhD for somebody else, you are always going to be doing it for yourself. If the only reason you are doing a PhD is because someone else told you that you should, then you are likely to have a really hard time. By the time you graduate, you may have a PhD but without any idea of what you're going to do with it.

@@

If you are planning a career outside academia, it is worth contacting potential employers and finding out how interested they are in employing people with a PhD in the subject area you are considering. Similarly, if you are considering a career inside academia, it is worth talking to people who are already in that job market (probably post-docs in your own group) about how easy or difficult they are finding it. Whatever you are planning to do, the right time for planning is before you begin your PhD. There is more written about planning your future [here](#Plan).

@@

## Your mental health will likely suffer
Doing a PhD will be the cause of stress not only in your working life, but will also impact other areas of your life. Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a [support network](#support). It is worth tracking your mental health during your PhD to check that you are not getting into difficulties, and I show you later [how to do this](#healthy).

@@

## What to avoid in an advisor
Remember that relationships occur on a one-to-one basis, and it may be that not every advisor gets on with every student. There are some advisors who have many problems with many students, and these are best avoided no matter how high their profile is [see @malaga2004meyer;@abbott2004junior; but also see @vences2004meyer for alternative viewpoints]. There are other advisors who are so busy that they are never able to meet with their students, and pass this task on to somebody else in their group. Again no matter how high profile these people are, if they can't advise their own students they are best avoided. Chamberlain [-@chamberlain2016ten] suggests that there are 10 types of academic advisors, implying that 9 of these are rather dysfunctional.

@@

It is possible that your potential advisor will tell you that they don't have any funding for a project that you may want to do. This is both a curse and an opportunity. There are many opportunities to raise money to fund PhD projects. Raising the funding yourself is one way of making sure it really remains your project. But the availability of this kind of funding does vary from country to country and you may want to embark on the fundraising side of your PhD before you register. Especially if you have a limited amount of time once you have registered in order to complete your PhD studies.

Some projects are much cheaper than others. For example, if your intention is to sequence genomes of every animal or plant that you work with, the finance required will likely be far greater than your bursary. If however your intention is to do a PhD that models distributions of populations of animals and plants, then the entire costs for your study may only add up to a computer, remote access to a computing cluster and data storage.

@@

## Core skills {#coreskills}

Although this book is about writing your PhD in Biological Sciences, there are a number of core skills that you, or anyone doing a PhD will need. This section of the book aims to give you a foundation on these core skills and encourage you to use them from the very start of your PhD studies. You may well find that you already have one or more of these core skills to a certain degree, but I would encourage you to read the chapters in this section and reflect on your proficiency of these core skills and build them up where you need to.

 - [Time management](#time)
 - [Planning](#plan)
 - [Communication](#email)
 - [Upskilling](#useR)
 - [Mental health](#healthy)
 - [Advisor relations](#advisor)
 - [Critical Reading](#critical)
 - [Regular writing](#regularwriting)
 - [A life scientific](#lifescientific)

Improving Each of these core skills will make doing your PhD easier. These core skills will extend beyond your PhD into your working life beyond such that you can consider them to be life-skills. None of them are independent, instead they are all integrated with each other and require careful thought and reflection from you about how you can improve each of them.
